Why do people punish? Evidence for a range of strategic concerns

### 2023-12-13 manuscript.tex
update manuscript

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{punishment; cooperation; economic games\newline\indent Word count: 5185 words}
\usepackage{lineno}

@@

\abstract{%
Costly punishment is thought to be one of the key mechanisms sustaining cooperation in humans. However, the motives for punitive behaviour remain unclear. Punishment is often assumed to be motivated by a desire to convert free riders into cooperators, but it is also consistent with a host of other functions, such as levelling payoffs or increasing one's relative position. We used a suite of six economic games to tease apart the different motives for punishment. Across representative samples from the United Kingdom and the United States, we estimated the frequency of different punishment strategies in the population, finding that egalitarian motives for punishment are more common than behaviour-change motives. Moreover, different punishment strategies were differentially predicted by personality, social preferences, political ideology, and religious views. Self-reports of behaviour in the games suggested that people have some degree of insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment.
}

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost\textsuperscript{1},
ostensibly encouraging cooperative behaviour from the target (or bystanders\textsuperscript{2--4}) in the future. Punishment
therefore offers a route to maintaining or increasing cooperation by changing
the payoff structure of social interactions such that it no longer pays to cheat
or exploit social partners\textsuperscript{1,5}.

In humans, many studies of punishment have been carried out in laboratory
settings using economic games\textsuperscript{6--15}. In these games, participants are usually given a sum of money
that they can use to invest in collective action or to help others.
Alternatively, participants can `cheat' by keeping the money for themselves or
by exploiting the contributions of others. Punishment is introduced into such
games by giving participants the option to pay a small `fee' to impose a greater
`fine' on their co-players. Several lines of experimental evidence indicate that
people use this punishment option\textsuperscript{12}, that they enjoy
punishing\textsuperscript{16}, and that they frequently, though not
always\textsuperscript{17}, punish cheating or exploitative
co-players\textsuperscript{10,11}.

Evidence from these experiments suggests that the threat of costly punishment
plays an important role in promoting human cooperation. People tend to cooperate
more in games where punishment is possible compared to those where it is not\textsuperscript{6,7,15}. The effect that the threat of
punishment has on cooperation is also evident in the higher contributions
typically observed in the Ultimatum Game (where punishment is possible) compared
to the structurally-similar Dictator Game (where it is not)\textsuperscript{18}. This
typical cooperation-enhancing effect of punishment has also been observed across
societies\textsuperscript{7}, leading some to suggest that costly punishment has
played a key role in the cultural evolution of cooperation in humans\textsuperscript{19--22}.

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to
achieve other ends. Some have argued that punishment is primarily used to shape
future behaviour, either to deter personal harm\textsuperscript{3,9,23} or to uphold normative standards of cooperative behaviour\textsuperscript{20,21,24--27}.
But while the \emph{threat} of punishment can have a cooperation-enhancing effect,
the \emph{enactment} of this punishment does not consistently deter targets from
cheating in the future\textsuperscript{15}. This calls into question whether
punishment primarily operates as a behaviour-change tool or whether it is used
to achieve other goals.

Beyond behaviour-shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused\textsuperscript{28}. Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others\textsuperscript{15,29}) and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs\textsuperscript{30}). Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position\textsuperscript{15}.

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games\textsuperscript{12,29,31,32}, on the very last round of repeated games\textsuperscript{33}, or in games where the target never learns about the
punishment\textsuperscript{34}; punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment\textsuperscript{17,35}); punishing in scenarios where they were not
personally harmed (third-party punishment\textsuperscript{36}); and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions\textsuperscript{30,37,38}.

The general conclusion from this research is that there is no one unifying
function of costly punishment in humans. Instead, punishment should be thought
of as a flexible behavioural tool that serves a variety of functions that are
not mutually exclusive\textsuperscript{15}. Due to its multipurpose nature, we should
therefore expect variation in punishment strategies in the population, much like
the observed variation in social learning strategies\textsuperscript{39}. Some
individuals may use punishment as a behaviour shaping tool, for example, while
others may use it to reduce or reverse payoff differentials.

@@

Here, we aim to delineate nine possible punishment strategies by asking
whether people punish in a manner consistent with a specific strategy and, if
so, what other characteristics (personality, social preferences, political
orientation) predict the use of different punishment strategies. Table
\ref{tab:tableStrategies} summarises the potential functions for costly
punishment in the economic games that we considered, and the behavioural
strategies they predict. Note that Table \ref{tab:tableStrategies} is not
an exhaustive list of all possible punishment strategies: we do not include
reputational functions of punishment in this table, such as signalling
trustworthiness\textsuperscript{4,43--46},
because our focus is on punishment strategies in anonymous economic games without
reputational incentives (but see ref\textsuperscript{47}).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{images/games} \caption{\emph{Visual summary of the six economic games.} In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does
nothing. Player 1 is then given the option to punish by paying a certain amount
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off by stealing, (B) Player 2 creates equality by stealing, (C) the
computer ``decides'' whether Player 2 steals, (D) the fee-fine ratio is 1:1, (E)
Player 2 is better off by stealing, and (F) Player 2 steals instead from a
third-party.}\label{fig:plotGames}
\end{figure}

@@

The overall pattern of punitive behaviour in the six economic games was in line
with previous research and very similar across both countries (Figure
\ref{fig:plotDecisions}). Participants were generally more likely to punish
targets who stole from another individual compared to targets who did not steal
(multilevel logistic regression; \emph{b} =
1.93, standard error =
0.27, \emph{p}
\textless{} .001). Participants were also more likely
to punish when targets' stealing behaviour generated inequalities,
specifically in Games E and F (\emph{b} =
2.42, SE = 0.44,
\emph{p} \textless{} .001).

@@

We classified participants into a particular strategy if their behaviour across
all twelve decisions matched our behavioural predictions shown in Table
\ref{tab:tableStrategies} exactly. Table \ref{tab:tableStrategyCounts} shows
the proportion of participants following each strategy, with N/A used to
represent participants who did not fit exactly into any particular strategy type.
Overall, 59\%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian).
Less common were the behaviour-shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek AI, competitive).
Although participants often punished targets who did not steal
in the six games (Figure \ref{fig:plotDecisions}), no participants followed the
antisocial strategy by exclusively punishing targets who did not steal across
\emph{all} games.

@@

\caption{\label{tab:tableStrategyCounts}Counts and proportions of participants
following each punishment strategy exactly, split by country. N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies.}

@@

The posterior estimates from the model are presented in Figure
\ref{fig:plotModel1b}. The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek AI strategies were the next most likely, with higher
posterior estimates than the competitive and antisocial strategies. None of the
other strategies differed in their posterior estimates. The same general pattern
emerged when we analysed the full dataset without pre-registered exclusions
(Supplementary Figure \ref{fig:plotModel1a}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotModel1b-1.pdf}
\caption{\label{fig:plotModel1b}\emph{Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model.} The model
assumes an implementation error rate of 5\%. Points represent posterior medians,
line ranges represent 50\% and 95\% credible intervals.}
\end{figure}

@@

Conversely, personality and social preferences were linked to variation in
punishment strategies. When including the Big-6 personality dimensions and
Social Value Orientation (SVO) in the model, we found associations with the
egalitarian, never punish, and random choice strategies (Figure
\ref{fig:plotAllPers2}). Participants higher in SVO were more likely to follow
the egalitarian and the never punish strategies, while those with lower SVO
scores were more likely to enact the random choice strategy. The personality
dimensions of honesty-humility and openness to experience were both positively
associated with following the never punish strategy, while extraversion
negatively predicted this strategy. The effects were mostly similar across
countries, but occasionally differed: for example, in the United States,
but not in the United Kingdom, honesty-humility was positively associated with
following the egalitarian strategy and negatively associated with following
the random choice strategy. Overall, the same pattern of results emerged when
analysing the full dataset without exclusions (Supplementary Figure
\ref{fig:plotAllPers1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPers2-1.pdf}
\caption{\label{fig:plotAllPers2}\emph{Posterior slopes from Bayesian latent state models
including Big-6 personality dimensions and Social Value Orientation.} Each row
represents a separate model. All models assume an implementation error rate of
5\%. Points represent posterior medians, line ranges represent 95\% credible
intervals.}
\end{figure}

Political and religious variables were also associated with punishment strategy
(Figure \ref{fig:plotAllPolRel2}). These effects tended to be more pronounced
in the United States. Controlling for Social Dominance Orientation, American
participants higher in Right Wing Authoritarianism were more likely to follow
the strategies avoiding disadvantageous inequity and seeking advantageous
inequity. Participants who stated that they would like to ``bring those below
them {[}on the socio-economic status ladder{]} up a peg'' were more likely to follow
the egalitarian strategy, while American participants higher in Social Dominance
Orientation, Right Wing Authoritarianism, and believing that God controls events
in the world were less likely to follow the egalitarian strategy. In general,
religious and conservative participants were less likely to follow the never
punish strategy. This general pattern of results was replicated with the full
dataset (Supplementary Figure \ref{fig:plotAllPolRel1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPolRel2-1.pdf}
\caption{\label{fig:plotAllPolRel2}\emph{Posterior slopes from Bayesian latent state models
including political ideology, views about social inequality, and religiosity.}
Each row represents a separate model aside from Social Dominance Orientation and
Right Wing Authoritarianism, which control for one another within the same
model. All models assume an implementation error rate of 5\%. Points represent
posterior medians, line ranges represent 95\% credible intervals.}
\end{figure}

@@

In general, we found that self-reported strategy usage was positively associated
with the behavioural strategy that participants employed (see Supplementary
Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2} for the distribution
of responses to self-report questions). Figure \ref{fig:plotAllSliders2} shows
the relationships between self-report questions and the different punishment
strategies, highlighting the combinations where the question matched the
behavioural strategy. We found positive relationships between the self-report
questions and strategy usage for the norm-enforcing, egalitarian, seek
advantageous inequity, never punish, and random choice strategies. The 95\%
credible intervals for other estimates included zero, though these estimates
often trended in a positive direction. The same pattern of results was found
when analysing the full dataset without exclusions (Supplementary Figure
\ref{fig:plotAllSliders1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllSliders2-1.pdf}
\caption{\label{fig:plotAllSliders2}\emph{Posterior slopes from models including
self-reported strategy usage.} Each row represents a separate model. All models
assume an implementation error rate of 5\%. Highlighted estimates represent
combinations where the self-report question matched the behavioural strategy.
Points represent posterior medians, line ranges represent 95\% credible
intervals.}
\end{figure}

@@

Using a suite of economic games measuring punishment in different situations,
we have shown that punishment does not serve just one function, but instead is
a flexible tool that can be and is used for different purposes\textsuperscript{15}.
Punishment is more akin to a swiss army knife than a hammer, used by some to
enforce norms of cooperation and by others to reduce or even create inequality
between individuals. We found that people's punishment strategy can, to some
extent, be predicted by individual differences in personality, social
preferences, and political and religious views. Moreover, contrary to the view
that people are often unable to articulate the reasons for their punitive
behaviour\textsuperscript{28}, people seem to have some degree of insight into the
strategy they are using. Despite small differences, these general patterns
replicated in samples from both the United Kingdom and the United States.

Among the punitive strategies, the most common were particularly sensitive to
inequality in payoffs, either from a self-referential perspective (i.e., avoid
disadvantageous inequity) or more generally (i.e., egalitarian). This is in
line with previous studies which have highlighted inequity aversion as an
important motivation for punishment in economic games\textsuperscript{29,30}. Personality and social preference variables mapped
onto these strategies in expected ways. Traits associated with other-regarding
concern, such as SVO and honesty-humility, predicted following the egalitarian
strategy, whereas religious and conservative individuals were less likely to
follow this strategy, especially in the United States. Moreover, participants
following the egalitarian strategy were able to self-report this strategy,
though the same was not true for the avoid disadvantageous inequity strategy.

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less
common than strategies sensitive to inequality in our set of games. This was
reflected both in participants' elicited punishment behaviour (Figure
\ref{fig:plotModel1b}) and in their self-reports of their own strategy
(Supplementary Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2}).
Regarding the predictors of these strategies, we found that demographic,
personality, political, and religious variables tended to be unrelated to
deterrent and norm-enforcing punishment strategies. We also found that
participants had insight into the norm-enforcing strategy, but not the deterrent
strategy. This finding is in line with previous research showing that people
struggle to accurately report the deterrent motivations for their punitive
behaviour\textsuperscript{28}.

Other punitive strategies were less common in our dataset, but some were more
prevalent than others. For example, participants were more likely to use
punishment to seek advantageous inequity than to exclusively harm those who did
not steal (i.e., antisocial punishment). The existence of the ``seek AI'' strategy
in our dataset supports the claim that punishment can also be used as a tool to
increase one's own relative position\textsuperscript{15}. While generally rare, we
found that this motive for punishment was more common among authoritarian
participants, at least in the United States, potentially providing an
explanation for why peer punishment has been found to be more common among
conservatives in previous work\textsuperscript{41}. Moreover, the fact that no
participants in our sample punished non-stealing across all games suggests that
antisocial punishment does not function to harm cooperators specifically, as has
been previously suggested\textsuperscript{17}. Instead, antisocial punishment appears
to be motivated by improving one's relative position in general, which is in
line with work showing that antisocial punishment disappears with a 1:1 fee-fine
ratio\textsuperscript{35}.

The fact that people use punishment for many different reasons poses problems
for the way that punishment is operationalised in classic behavioural economic
game studies. In these studies, a common assumption is that participants will
punish to change the behaviour of cheats. But in reality, people may be choosing
the punishment option to achieve a variety of different goals. The targets of
punishment in these studies are likely well aware that punishment could be
levied for these different reasons and this knowledge may impact their
responses. For example, if cheating targets interpret punishment as serving a
competitive motive, it may elicit retaliation rather than encourage
cooperation\textsuperscript{9,15,50}. This might help to
explain the mixed findings in the field as to whether punishment actually
motivates cheating targets to cooperate in the future\textsuperscript{15}.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{punishment; cooperation; economic games\newline\indent Word count: xxxx words}
\usepackage{lineno}

@@

\abstract{%
Costly punishment is thought to be a key mechanism sustaining human cooperation. However, the motives for punitive behaviour remain unclear. Although often assumed to be motivated by a desire to convert cheats into cooperators, punishment is also consistent with other functions, such as levelling payoffs or improving one's relative position. We used six economic games to tease apart different motives for punishment and to explore whether different punishment strategies were associated with personality variables, political ideology, and religiosity. We used representative samples from the United Kingdom and the United States (N = 2010) to estimate the frequency of different punishment strategies in the population. The most common strategy was to never punish: prosocial individuals and those scoring high in honesty-humility were among the least punitive in our sample, while religious individuals and those scoring high for right-wing authoritarianism and social dominance orientation were more punitive. For people who did punish, strategy use was more consistent with egalitarian motives than behaviour-change motives. Nevertheless, different punishment strategies were also associated with personality, social preferences, political ideology, and religiosity. Self-reports of behaviour in the games suggested that people have some insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment and show how the different motives underpinning punishment decisions are linked with core character traits.
}

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost\textsuperscript{1},
ostensibly encouraging cooperative behaviour from the target (or bystanders\textsuperscript{2--4}) in the future. Punishment offers a
route to maintaining or increasing cooperation by changing the payoff structure
of social interactions such that it no longer pays to cheat or exploit social
partners\textsuperscript{1,5}. Yet, despite its theoretical
importance, the question of why people choose to punish others is still hotly
contested\textsuperscript{6}. In this study, we use a battery of economic games to
disentangle the different motives underpinning punishment and explore how these
motives vary across individuals.

In humans, many studies of punishment have been carried out in laboratory
settings using economic games\textsuperscript{6--15}. In these games, participants are usually given a sum of money
that they can use to invest in collective action or to help others.
Alternatively, participants can `cheat' by keeping the money for themselves or
by exploiting the contributions of others. Punishment is introduced into such
games by giving participants the option to pay a small `fee' to impose a greater
`fine' on their co-players. Several lines of experimental evidence indicate that
people use this punishment option\textsuperscript{13}, that they enjoy
punishing\textsuperscript{16}, and that they frequently, though not
always\textsuperscript{17}, punish cheating or exploitative
co-players\textsuperscript{11,12}.

Evidence from these experiments suggests that the threat of costly punishment
plays an important role in promoting human cooperation. People tend to cooperate
more in games where punishment is possible compared to those where it is not\textsuperscript{6--8}. The effect that the threat of
punishment has on cooperation is also evident in the higher contributions
typically observed in the Ultimatum Game (where punishment is possible) compared
to the structurally-similar Dictator Game (where it is not)\textsuperscript{18}. This
typical cooperation-enhancing effect of punishment has also been observed across
societies\textsuperscript{8}, leading some to suggest that costly punishment has
played a key role in the cultural evolution of cooperation in humans\textsuperscript{19--22}.

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to
achieve other ends. Some have argued that punishment is primarily used to shape
future behaviour, either to deter personal harm\textsuperscript{3,10,23} or to uphold normative standards of cooperative behaviour\textsuperscript{20,21,24--27}.
But while the \emph{threat} of punishment can have a cooperation-enhancing effect,
the \emph{enactment} of this punishment does not consistently deter targets from
cheating in the future\textsuperscript{6}. This calls into question whether
punishment primarily operates as a behaviour-change tool or whether it is used
to achieve other goals.

Beyond behaviour shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused\textsuperscript{28}. Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others\textsuperscript{6,29}) and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs\textsuperscript{30}). Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position\textsuperscript{6}.

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games\textsuperscript{13,29,31,32}, on the very last round of repeated games\textsuperscript{33}, or in games where the target never learns about the
punishment\textsuperscript{34}; punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment\textsuperscript{17,35}); punishing in scenarios where they were not
personally harmed (third-party punishment\textsuperscript{36}); and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions\textsuperscript{30,37,38}.

The general conclusion from this research is that there is no one unifying
function of costly punishment in humans. Instead, punishment should be thought
of as a flexible behavioural tool that serves a variety of functions that are
not mutually exclusive\textsuperscript{6}. Due to its multipurpose nature, we should
therefore expect variation in punishment strategies in the population, much like
the observed variation in social learning strategies\textsuperscript{39}. Some
individuals may use punishment as a behaviour shaping tool, for example, while
others may use it to reduce or reverse payoff differentials.

@@

Here, we aim to delineate nine possible punishment strategies by asking
whether people punish in a manner consistent with a specific strategy and, if
so, what other characteristics (personality, social preferences, political
orientation, religious views) predict the use of different punishment strategies.
Table \ref{tab:tableStrategies} summarises the potential functions for costly
punishment in the economic games that we considered, and the behavioural
strategies they predict. Note that Table \ref{tab:tableStrategies} is not
an exhaustive list of all possible punishment strategies: we do not include
reputational functions of punishment in this table, such as signalling
trustworthiness\textsuperscript{4,43--46},
because our focus is on punishment strategies in anonymous economic games without
reputational incentives (but see ref\textsuperscript{47}).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{images/games} \caption{\emph{Visual summary of the six economic games.} In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does
nothing. Player 1 is then given the option to punish by paying a certain amount
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C)
the computer ``decides'' whether Player 2 steals, (D) the fee-fine ratio is 1:1,
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from
a third-party.}\label{fig:plotGames}
\end{figure}

@@

The overall pattern of punitive behaviour in the six economic games was very
similar across both countries (Figure \ref{fig:plotDecisions}). Participants
were generally more likely to punish targets who stole compared to targets who
did not steal (multilevel logistic regression; \emph{b} =
1.93, standard error =
0.27, \emph{p}
\textless{} .001). Participants were also more likely
to punish when targets' stealing behaviour generated inequalities,
specifically in Games E and F (\emph{b} =
2.42, SE = 0.44,
\emph{p} \textless{} .001).

@@

We classified participants into a particular strategy if their behaviour across
all twelve decisions matched our behavioural predictions shown in Table
\ref{tab:tableStrategies} exactly. Table \ref{tab:tableStrategyCounts} shows
the proportion of participants following each strategy, with N/A used to
represent participants who did not fit exactly into any particular strategy type.
Overall, 59\%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian).
Less common were the behaviour shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek advantageous
inequity, competitive). Although participants often punished targets who did not
steal in the six games (Figure \ref{fig:plotDecisions}), no participants
followed the antisocial strategy by exclusively punishing targets who did not
steal across \emph{all} games.

@@

\caption{\label{tab:tableStrategyCounts}Counts and proportions of participants
following each punishment strategy exactly, split by country. \emph{N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies. AI = advantageous inequity, DI = disadvantageous inequity.}}

@@

The posterior estimates from the model are presented in Figure
\ref{fig:plotModel1b}. The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek advantageous inequity strategies were the next most
likely, with higher posterior estimates than the competitive and antisocial
strategies. None of the other strategies differed in their posterior estimates.
The same general pattern emerged when we analysed the full dataset without
pre-registered exclusions (Supplementary Figure \ref{fig:plotModel1a}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotModel1b-1.pdf}
\caption{\label{fig:plotModel1b}\emph{Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model.} The model
assumes an implementation error rate of 5\%. Points represent posterior medians,
line ranges represent 50\% and 95\% credible intervals. AI = advantageous
inequity, DI = disadvantageous inequity.}
\end{figure}

@@

Conversely, personality and social preferences were linked to variation in
punishment strategies. When including the Big-6 personality dimensions and
Social Value Orientation (SVO) in the model, we found associations with the
egalitarian, never punish, and random choice strategies, with small-to-medium
effect sizes (Figure \ref{fig:plotAllPers2}). More prosocial participants
(those with higher SVO scores) were more likely to follow the egalitarian and
the never punish strategies, while those with lower SVO scores were more likely
to enact the random choice strategy. The personality dimensions of
honesty-humility and openness to experience were both positively associated with
following the never punish strategy, while extraversion negatively predicted
this strategy. The effects were mostly similar across countries, but
occasionally differed: for example, in the United States, but not in the United
Kingdom, honesty-humility was positively associated with following the
egalitarian strategy and negatively associated with following the random
choice strategy. Overall, the same pattern of results emerged when analysing the
full dataset without exclusions (Supplementary Figure \ref{fig:plotAllPers1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPers2-1.pdf}
\caption{\label{fig:plotAllPers2}\emph{Posterior slopes from Bayesian latent state models
including Big-6 personality dimensions and Social Value Orientation.} Each row
represents a separate model. All models assume an implementation error rate of
5\%. Points represent posterior medians, line ranges represent 95\% credible
intervals. AI = advantageous inequity, DI = disadvantageous inequity.}
\end{figure}

Political and religious variables were also associated with punishment strategy
(Figure \ref{fig:plotAllPolRel2}). These effects were small-to-medium in size
and tended to be more pronounced in the United States. Controlling for Social
Dominance Orientation, American participants higher in Right Wing
Authoritarianism were more likely to follow the strategies avoiding
disadvantageous inequity and seeking advantageous inequity. Participants who
stated that they would like to ``bring those below them {[}on the socio-economic
status ladder{]} up a peg'' were more likely to follow the egalitarian strategy,
while American participants higher in Social Dominance Orientation, Right Wing
Authoritarianism, and believing that God controls events in the world were less
likely to follow the egalitarian strategy. In general, religious and
conservative participants were less likely to follow the never punish strategy.
This general pattern of results was replicated with the full dataset
(Supplementary Figure \ref{fig:plotAllPolRel1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPolRel2-1.pdf}
\caption{\label{fig:plotAllPolRel2}\emph{Posterior slopes from Bayesian latent state models
including political ideology, views about social inequality, and religiosity.}
Each row represents a separate model aside from Social Dominance Orientation and
Right Wing Authoritarianism, which control for one another within the same
model. All models assume an implementation error rate of 5\%. Points represent
posterior medians, line ranges represent 95\% credible intervals. AI =
advantageous inequity, DI = disadvantageous inequity.}
\end{figure}

@@

In general, we found that self-reported strategy usage was positively associated
with the behavioural strategy that participants employed, with effects ranging
from small to large in size (see Supplementary Figures \ref{fig:plotSliders1}
and \ref{fig:plotSliders2} for the distribution of responses to self-report
questions). Figure \ref{fig:plotAllSliders2} shows the relationships between
self-report questions and the different punishment strategies, highlighting the
combinations where the question matched the behavioural strategy. We found
positive relationships between the self-report questions and strategy usage for
the norm-enforcing, egalitarian, seek advantageous inequity, never punish, and
random choice strategies. The 95\% credible intervals for other estimates
included zero, though these estimates often trended in a positive direction. The
same pattern of results was found when analysing the full dataset without
exclusions (Supplementary Figure \ref{fig:plotAllSliders1}).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllSliders2-1.pdf}
\caption{\label{fig:plotAllSliders2}\emph{Posterior slopes from models including
self-reported strategy usage.} Each row represents a separate model. All models
assume an implementation error rate of 5\%. Highlighted estimates represent
combinations where the self-report question matched the behavioural strategy.
Points represent posterior medians, line ranges represent 95\% credible
intervals. AI = advantageous inequity, DI = disadvantageous inequity.}
\end{figure}

@@

Using a suite of economic games measuring punishment in different situations,
we have shown that punishment does not serve just one function, but instead is
a flexible tool that can be and is used for different purposes\textsuperscript{6}.
While some use punishment to enforce norms of cooperation, others use it to
reduce or even create inequality between individuals. We found that people's
punishment strategy can, to some extent, be predicted by individual differences
in personality, social preferences, and political and religious views. Moreover,
contrary to the view that people are often unable to articulate the reasons for
their punitive behaviour\textsuperscript{28}, people seem to have some degree of
insight into the strategy they are using. Despite small differences, these
general patterns replicated in samples from both the United Kingdom and the
United States, providing further confidence in the results.

Among the punitive strategies, the most common were particularly sensitive to
inequality in payoffs, either from a self-referential perspective (i.e., avoid
disadvantageous inequity) or more generally (i.e., egalitarian). This is in
line with previous studies which have highlighted inequity aversion as an
important motivation for punishment in economic games\textsuperscript{29--31,37,48,50}.
Personality, social preferences, and political and religious views were related
to these strategies. Traits associated with other-regarding concern, such as SVO
and honesty-humility, predicted following the egalitarian strategy. By contrast,
religious and conservative individuals, including individuals higher in SDO and
RWA, were less likely to follow the egalitarian strategy and more likely to
follow the avoid disadvantageous inequity strategy, especially in the United
States. Authoritarian participants in the United States were also more likely to
actively seek advantageous inequity with their punishment. These findings
provide insight into the motives that may have driven higher levels of peer
punishment among conservatives in previous work\textsuperscript{41}.

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less
common than strategies sensitive to inequality in our set of games. This was
reflected both in participants' elicited punishment behaviour (Figure
\ref{fig:plotModel1b}) and in their self-reports of their own strategy
(Supplementary Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2}).
Although our design did not explicitly allow for behaviour shaping as the
interactions were all one-shot, we did manipulate whether the target's stealing
behaviour was intentional or not (Game C), an approach which has been used in
previous vignette studies to identify behaviour shaping motives\textsuperscript{28}.
The lower prevalence of behaviour shaping strategies in our study is consistent
with prior work showing that punishment often continues to be used when
behaviour shaping is impossible, such as when the target will never find out
that they have been punished\textsuperscript{34} or on the last round of repeated
interactions\textsuperscript{33}. We found that demographic, personality, political,
and religious variables tended to be unrelated to behaviour shaping strategies,
though this could reflect low power to detect such associations given the low
prevalence of these strategies in our sample. We also found that participants
accurately reported using the norm-enforcing strategy, but not the deterrent
strategy. This finding is in line with previous research showing that people
struggle to accurately report the deterrent motivations for their punitive
behaviour\textsuperscript{28}.

We defined antisocial punishment as occurring when a participant exclusively
punished those who did not steal. There were no participants in our sample who
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all
games suggests that the antisocial punishment that has been observed in other
studies\textsuperscript{17,51} is not aimed at harming cooperators
specifically, as has been previously suggested\textsuperscript{17}. Instead,
antisocial punishment is more likely to be motivated by improving one's
relative position, which is in line with work showing that antisocial punishment
disappears when relative payoffs cannot be changed (e.g., when punishment is
only available with a 1:1 fee-fine ratio\textsuperscript{35}).

The fact that people use punishment for many different reasons poses problems
for the way that punishment is operationalised in classic behavioural economic
game studies. In these studies, a common assumption is that participants punish
to change the behaviour of cheats\textsuperscript{11,12}. But in reality, people
may be choosing the punishment option to achieve a variety of different goals.
This has implications for how people respond to being targeted by punishers in
these games. Targets of punishment in these studies may know that punishment
reflects different motives and can respond accordingly. For example, if targets
interpret punishment as serving a competitive motive, it may elicit retaliation
rather than encourage cooperation\textsuperscript{6,10,52}.
As punishers' motives must be inferred (and such inferences likely depend on
character traits of the target, as well as the context in which punishment
occurs), there is likely to be some variation and error in attributing motives
to punishers. To the extent that inferred motives affect target responses, this
might help to explain the mixed findings in the field as to whether punishment
actually motivates cheating targets to cooperate in the future\textsuperscript{6}.
